---
title: "Linear regression: Bayesian version in JAGS, credible and prediction intervals"
author: "Petr Keil"
date: "October 2023"
output:
  html_document:
    highlight: pygments
    keep_md: yes
    number_sections: yes
    theme: cerulean
    toc: yes
  pdf_document: default
---

***

# Objective

Here we will do classical linear regression in a Bayesian setting. We will also show the difference between **credible intervals** of parameters and  **prediction intervals** of predictions.

In **Part 1** the data and the model are introduced. Participants then try to implement the model in BUGS. In **Part 2** the solution is exposed, and the difference between credible and prediction intervals is explained.

***

# The data

```{r}
  catepil <- read.csv("https://rawgit.com/petrkeil/ML_and_Bayes_2017_iDiv/master/Linear%20Regression/catepilar_data.csv")
  catepil
```

# Preparing the data for JAGS

We will put all of our data in a list object

```{r}
linreg.data <- list(N=9,
                    tannin=catepil$tannin,
                    growth=catepil$growth)
linreg.data

```

# Fitting the model by MCMC in JAGS

The R library that connects R with JAGS:

```{r, fig.width=4, fig.height=4, tidy=FALSE, message=FALSE, warning=FALSE}
library(R2jags)
```

This is the JAGS definition of the model:

```{r, tidy=FALSE, message=FALSE, warning=FALSE}
cat("
  model
  {
    # priors
    a ~ dnorm(0, 0.001) # intercept
    b ~ dnorm(0, 0.001) # slope
    sigma ~ dunif(0, 100) # SD 
  
    # JAGS is strange, it uses precision instead of SD
    tau <- 1/(sigma*sigma) # precision
    
    # likelihood
    for(i in 1:N)
    {
      mu[i] <- a + b*tannin[i]
      growth[i] ~ dnorm(mu[i], tau)
    }
  }
", file="linreg_model.bug")
```

The MCMC sampling done by ```jags``` function:

```{r, tidy=FALSE, message=FALSE, warning=FALSE}
model.fit <- jags(data=linreg.data, 
               model.file="linreg_model.bug",
               parameters.to.save=c("a", "b", "sigma"),
               n.chains=3,
               n.iter=2000,
               n.burnin=1000)
```

And we can examine the results:

```{r, fig.width=8, fig.height=12, eval=FALSE}
model.fit
plot(as.mcmc(model.fit))
```

# Bayesian predictions using MCMC

**Beware:** Here we will use library ```rjags``` that gives more control over the MCMC (as opposed to ```R2jags```):

```{r, fig.width=4, fig.height=4, tidy=FALSE, message=FALSE, warning=FALSE}
library(rjags)

cat("
  model
  {
    # priors
    a ~ dnorm(0, 0.001) # intercept
    b ~ dnorm(0, 0.001) # slope
    sigma ~ dunif(0, 100) # SD 
  
    # JAGS is strange, it uses precision instead of SD
    tau <- 1/(sigma*sigma) # precision
    
    # likelihood
    for(i in 1:N)
    {
      mu[i] <- a + b*tannin[i]
      growth[i] ~ dnorm(mu[i], tau)
    }
  
    # predictions
    for(i in 1:N)
    {
      prediction[i] ~ dnorm(mu[i], tau) 
    }
  }
", file="linreg_model.bug")
```

Initializing the model:

```{r, fig.width=4, fig.height=4, tidy=FALSE, message=FALSE, warning=FALSE}
jm <- jags.model(data=linreg.data, 
                 file="linreg_model.bug",
                 n.chains = 3, n.adapt=1000)
```

The burn-in phase:

```{r}
update(jm, n.iter=1000)
```

We will monitor the predicted values:

```{r}
params <- c("prediction")
samps <- coda.samples(jm, params, n.iter=5000)     
predictions <- summary(samps)$quantiles
```

And here we will monitor the expected values:

```{r}
params <- c("mu")
samps <- coda.samples(jm, params, n.iter=5000)     
mu <- summary(samps)$quantiles
```

Plotting the predictions:

```{r, fig.width=5, fig.height=5}
plot(c(0,8), c(0,18), type="n", xlab="tannin", ylab="growth")
points(catepil$tannin, catepil$growth, pch=19)
lines(catepil$tannin, mu[,"50%"], col="red")
lines(catepil$tannin, mu[,"2.5%"], col="red", lty=2)
lines(catepil$tannin, mu[,"97.5%"], col="red", lty=2)
lines(catepil$tannin, predictions[,"2.5%"], lty=3)
lines(catepil$tannin, predictions[,"97.5%"], lty=3)
```

The figure shows the **median expected value** (solid red), **95% credible intervals of the expected value** (dashed red) and **95% prediction intervals** (dotted). 



